---
title: "P8106 Midterm Final Report"
author:
- "Ila Kanneboyina, Shayne Estill, Naomi Simon-Kumar (ns3782)"
date: "03/24/2025"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}

# Ensures R code is suppressed
knitr::opts_chunk$set(
  echo = FALSE,         
  warning = FALSE,      
  message = FALSE,      
  results = 'hide'      
)

```

```{r}

## DATA CLEANING AND PREPARATION

# Load training data
load("dat1.RData") 
dat1 <- dat1

# Load test data (dat2)
load("dat2.RData") 
dat2 <- dat2

# Set seed for reproducibility
set.seed(299)

data_split <- initial_split(dat1, prop = 0.8)

# Set 10-fold cross-validation
ctrl1 <- trainControl(method = "cv", number = 10)

# Remove non-predictor variables
training_data <- training(data_split) %>% select(-id) # remove ID variable 
testing_data <- testing(data_split) %>% select(-id) # remove ID variable 
dat2 <- dat2 %>% select(-id) # remove ID variable 

# Convert categorical variables to factor: Training Data
training_data <- training_data %>%
  mutate(gender = factor(gender, 
                         levels = c(0, 1), 
                         labels = c("Female", "Male")), 
    race = factor(as.character(race), 
                  levels = c("1", "2", "3", "4"), 
                  labels = c("White", "Asian", "Black", "Hispanic")),
    smoking = factor(as.character(smoking), levels = c("0", "1", "2"), 
                     labels = c("Never", "Former", "Current")),
    diabetes = factor(diabetes, 
                      levels = c(0, 1), 
                      labels = c("No", "Yes")),
    hypertension = factor(hypertension, 
                          levels = c(0, 1), 
                          labels = c("No", "Yes")))

# Convert categorical variables to factor: Testing Data
testing_data <- testing_data %>%
  mutate(gender = factor(gender, 
                         levels = c(0, 1), 
                         labels = c("Female", "Male")), 
    race = factor(as.character(race), 
                  levels = c("1", "2", "3", "4"), 
                  labels = c("White", "Asian", "Black", "Hispanic")),
    smoking = factor(as.character(smoking), levels = c("0", "1", "2"), 
                     labels = c("Never", "Former", "Current")),
    diabetes = factor(diabetes, 
                      levels = c(0, 1), 
                      labels = c("No", "Yes")),
    hypertension = factor(hypertension, 
                          levels = c(0, 1), 
                          labels = c("No", "Yes")))

# Convert categorical variables to factor: Dat2

dat2 <- dat2 %>%
  mutate(gender = factor(gender, 
                         levels = c(0, 1), 
                         labels = c("Female", "Male")), 
    race = factor(as.character(race), 
                  levels = c("1", "2", "3", "4"), 
                  labels = c("White", "Asian", "Black", "Hispanic")),
    smoking = factor(as.character(smoking), levels = c("0", "1", "2"), 
                     labels = c("Never", "Former", "Current")),
    diabetes = factor(diabetes, 
                      levels = c(0, 1), 
                      labels = c("No", "Yes")),
    hypertension = factor(hypertension, 
                          levels = c(0, 1), 
                          labels = c("No", "Yes")))

```

# 1. Introduction

## 1.1. Background and Study Objective

This study aims to build a prediction model for antibody responses to a newly authorized vaccine, as measured by log-transformed antibody levels from dried blood spot samples. Our objective is to develop an accurate model to improve understanding of vaccine effectiveness across different population segments and to support the monitoring of immune protection over time.

## 1.2. Data Source and Description

The primary dataset (**dat1.RData**) contains demographic and clinical information from participants in a vaccine response study. Variables include age, gender, race/ethnicity, body mass index (BMI), blood pressure, cholesterol levels, diabetes, hypertension, and time since vaccination (in days). The outcome of interest is the log-transformed antibody level, measured using dried blood spot samples. The dataset consists of records from 5000 participants with 13 variables, excluding the unique identifier variable.
A second independent dataset (**dat2.RData**) with identical structure was collected several months later to assess model generalizability. The second dataset consists of records from 1000 participants.

# 2. Exploratory Analysis

## 2.1. Dataset Overview

```{r}


```

## 2.2. Summary Statistics

```{r}


```

## 2.3. Correlation Analysis

```{r}

## Correlation Plot

# Matrix of predictors 
x <- model.matrix(log_antibody ~ ., training_data)[, -1]

# Vector of response
y <- training_data$log_antibody

# Produce corrplot
corrplot::corrplot(cor(x), method = 'circle', type = 'full')

```

## 2.4. Exploratory Plots

```{r}


```


# 3. Model Training

## 3.1. Data Preparation

```{r}


```

## 3.2. Cross-validation approach

For model training and evaluation, we implemented 10-fold cross-validation on the training dataset (80% of the original data) using the trainControl function This partitioned the training data into 10 equal subsets, where each model was trained on 9 folds and validated on the remaining fold, rotating through all folds.

## 3.3. Models

### 3.3.1. Linear Regression

```{r}

```

### 3.3.2. Elastic Net

```{r}

```

### 3.3.3. MARS

MARS was selected as a candidate model as its ability to automatically detect important predictors and their interactions made it suitable for our dataset’s combination of demographic and clinical variables.

For model tuning, we initially specified a grid with degrees 1 to 3 (maximum number of interactions) and 2 to 15 retained terms, then expanded to degrees 1 to 4 and 2 to 20 retained terms to evaluate whether higher-order interactions might better capture complex relationships between variables. Cross-validation results showed that models with degree = 1 consistently achieved the lowest RMSE. As interaction degree increased, RMSE values became more variable rather than decreasing steadily, suggesting no gain in prediction accuracy. The range of retained terms appeared appropriate as RMSE decreased initially with more terms but then plateaued. The final MARS model (**degree = 1, 10 retained terms**) was able to represent nonlinear relationships through hinge functions at specific threshold values.

```{r}

# Set seed for reproducibility
set.seed(299)

# Specify MARS grid - first attempt 
# mars_grid <- expand.grid(
#  degree = 1:3, # degree of interactions
#  nprune = 2:15  # no. of retained terms
# )

# MARS tuning grid - expanding grid 
mars_grid <- expand.grid(
  degree = 1:4,     # interaction degrees
  nprune = 2:20     # number of terms
)

# Train MARS model to predict log_antibody
mars.fit <- train(log_antibody ~ .,
                  data = training_data,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl1)

# Plot CV performance
ggplot(mars.fit)


# Optimal parameters
mars.fit$bestTune

# Final model coefficients
mars_coef <- coef(mars.fit$finalModel)

```

### 3.3.4. GAM

GAM was selected because immune responses typically follow smooth, nonlinear patterns that change gradually with predictors like age, BMI, and time since vaccination. GAM’s ability to model these relationships as flexible smooth functions while also being interpretable made it an appropriate candidate model to examine how each predictor independently contributes to antibody levels.
Our analysis showed that some predictors have estimated degrees of freedom (edf) greater than 1, specifically height (edf=3.163), weight (edf=1.255), BMI (edf=4.700), and time (edf=7.665), indicating they are modeled as nonlinear smooth functions. This is supported by the GAM predictor plots. Weight, while slightly above 1, was not statistically significant.
Among the nonlinear terms, height, BMI, and time were statistically significant (p < 0.05), indicating meaningful nonlinear relationships with the outcome.
Predictors age, SBP, and LDL each have an edf of 1.00, indicating that they are modeled as approximately linear. Of these, only age was statistically significant (p < 2e-16), suggesting a strong linear association with log antibody levels. 

```{r}

# Set seed for reproducibility
set.seed(299)

# Fit a GAM model, using training data
gam_antibody <- gam(log_antibody ~ gender + race + 
                      smoking + s(age) +
                      s(height) + s(weight) + s(bmi) +
                      diabetes + hypertension + s(SBP) + 
                      s(LDL) + s(time), 
                    data = training_data)

# Plot smooth GAM terms
# Code source: https://bookdown.org/ndphillips/YaRrr/arranging-plots-with-parmfrow-and-layout.html
par(mfrow = c(3, 2))  
plot(gam_antibody)
par(mfrow = c(1, 1)) # Reset plotting window


# Use train() from caret to fit GAM Model
gam.fit <- train(
  log_antibody ~ gender + race + 
    smoking + age +
    height + weight + bmi + diabetes + 
    hypertension + SBP + 
    LDL + time,
  data = training_data,
  method = "gam",
  trControl = ctrl1
)

# Predictor EDF values from mgcv fitted GAM model
summary(gam_antibody)

```

# 4. Results

## 4.1. Model Comparison and Selection


```{r}


```

## 4.2. Final Model: xxx

* Detailed specification of the selected model (coefficients, terms, etc.)
* Visualisations if appropriate

```{r}

```

### 4.2.1. Evaluation of model performance

Test set performance metrics & interpretation

```{r}

```

## 4.3. Model Generalizability

Evaluation of model based on dat2.

```{r}

```

# 5. Conclusion

```{r}

```


# 6. References

Phillips, N. D. (n.d.). YaRrr! The Pirate's Guide to R. Retrieved March 29, 2025, from https://bookdown.org/ndphillips/YaRrr/arranging-plots-with-parmfrow-and-layout.html

# 7. Appendices

If relevant

```{r}

```

